---
title: "stitches in R"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# IMPORTANT NOTES

- Necessary to avoid R variable names like `variable.name` since `.` in python is somewhat analogous to `$` in R. 
- Follow the instructions in the `stitches-in-R-setup` R markdown so that this notebook will work. If you've followed that markdown, `stitches` should be installed in the `r-reticulate` virtual environment for this notebook and should be callable. 

# Setup

## R
```{r}
#### R  BLOCK ####
library(reticulate)
# # indicate that we want to use a specific virtualenv
use_virtualenv("r-reticulate", required =TRUE)


library(dplyr)
library(tidyr)
library(ggplot2)

# time slice for the map plotting:
map_time <- '2100-12-31'
```

## python
```{python}
#### Python Block ####
import stitches 
import pandas as pd
import pkg_resources
import xarray as xr
import numpy as np
pd.set_option('display.max_columns', None)


# The CMIP6 ESM we want to emulate and the variables we want to
# emulate
esm = ['CanESM5']
vars1 = ['tas', 'pr', 'hurs', 'psl']
```

# Load Hector data
pulled a quick rcp4.5 from HectorUI
```{r}
#### R  BLOCK ####
hector_tgav <- read.csv('data/Hector-data-2022-09-22.csv', skip=3, stringsAsFactors = FALSE)
hector_exp <- unique(hector_tgav$scenario)
head(hector_tgav)
```


# Reshape the Hector data to be stitches target

First we reshape the Hector time series to match the CMIP6 ESM GSAT smooth anomaly time series that STITCHES operates on:
NOTE we should probably add this as a function to stitches 
NOTE Hector time series start in 1800 - there's no science reason we couldn't do matching over that window and have a gridded netcdf of an extra 50 years of history. 
```{r}
#### R  BLOCK ####
hector_tgav %>%
    mutate(variable = 'tas',
           ensemble = 'hectorUI1',# doesn't affect the matching or stitching
           model = 'Hector') %>%
    rename(experiment = scenario) %>% 
    filter(year >= 1850) %>%
    select(variable, experiment, ensemble, model, year, value,  -units) ->
    target_tgav

print(head(target_tgav))
```

# Convert Hector Tgav to stitches windows

Now that we have shaped it as stitches is expecting, we can use stitches functions to calculate the matching windows for this target data

```{python}
#### Python Block ####
# smooth out Hector's historic volcano stuff and
# save off the smoothed Hector time series we actually use for matching so can plot later:      
target_hector = stitches.fx_processing.calculate_rolling_mean(r.target_tgav, size =31)

target_data = stitches.fx_processing.get_chunk_info(
    stitches.fx_processing.chunk_ts(df = target_hector,  n=9))
      
print(target_data.head())
print(target_data.head())
```

# Load the stitches archive data for matching

```{python, echo=FALSE}
exps = ['ssp126', 'ssp245', 'ssp370', 'ssp460', 'ssp585']
# pangeo table of ESMs for reference
pangeo_path = pkg_resources.resource_filename('stitches', 'data/pangeo_table.csv')
pangeo_data = pd.read_csv(pangeo_path)
# print(np.sort(pangeo_data.variable.unique()))
pangeo_data = pangeo_data[((
                               (pangeo_data['variable'].isin(vars1)) )  )
                          & ((pangeo_data['domain'].str.contains('mon')) ) &
                           ((pangeo_data['experiment'].isin(exps))) &
                           (pangeo_data['model'].isin(esm))].copy()


pangeo_good_ensembles =[]

for name, group in pangeo_data.groupby(['model', 'experiment', 'ensemble']):
    df = group.drop_duplicates().copy()
    if len(df) >= len(vars1):
        pangeo_good_ensembles.append(df)
    del(df)
    
pangeo_good_ensembles1 = pd.concat(pangeo_good_ensembles)
pangeo_good_ensembles2  = pangeo_good_ensembles1[['model', 'experiment', 'ensemble']].drop_duplicates().copy()
pangeo_good_ensembles = pangeo_good_ensembles2.reset_index(drop=True).copy()
del(pangeo_good_ensembles1)
del(pangeo_good_ensembles2)

# load the full archive data and subset to above
path = pkg_resources.resource_filename('stitches', 'data/matching_archive.csv')
archive_data = pd.read_csv(path)


# Keep only the entries that appeared in pangeo_good_ensembles:
keys =['model', 'experiment', 'ensemble']
i1 = archive_data.set_index(keys).index
i2 = pangeo_good_ensembles.set_index(keys).index
archive_data= archive_data[i1.isin(i2)].copy()
del(i1)
del(i2)

print(archive_data.head())
```

# Match
```{python}
#### python block ####
my_recipes = stitches.make_recipe(target_data, archive_data,
                                  non_tas_variables=['pr', 'psl', 'hurs'],
                                  tol = 0.14, N_matches = 10000,
                                  reproducible = True)
print(my_recipes.head())
```

Unless otherwise specified by the argument `reproducible = True`, the draws stitches takes from the pool of matches is stochastic. 

Within each of these ensembles, there's no envelope collapse. If we were to concatenate them together into
a super-ensemble, there would be. So it's not as powerful as a very large ensemble of collapse free realizations but it's still a distinct ensemble to consider.



# stitch Tgav

```{python}
#### python block ####
stitched_global_temp = stitches.gmat_stitching(my_recipes)
print(stitched_global_temp.head())
```


# plot Tgav
```{r, echo = FALSE}
#### R block ####
stitched_tgav <- py$stitched_global_temp

# some reshaping to make plotting easier
stitched_tgav %>%
    mutate(tmp = stitching_id) %>%
    separate(tmp, into = c('scenario', 't1', 't2'), sep = '~') %>%
    select(-t1, -t2) ->
    stitched_tgav

ggplot(data = stitched_tgav, aes(x= year, y = value, color = stitching_id)) +
    geom_line(size=0.2) +
  geom_line(data = hector_tgav, aes(x = year, y=value), color='black', size = 1.1) + 
  geom_line(data = py$target_hector, aes(x = year, y=value), color='blue', size = 0.7) + 
  ylab('degC') +
  ggtitle('global average temperature anomaly - Ensemble 1')+ 
  theme(legend.position="bottom") +
  theme(legend.title = element_text( size=4), legend.text=element_text(size=4))
```



# stitch gridded
Using the above recipes, stitches can produce new gridded netcdf files for each recipe, for multiple variables. 
Creating multiple netcdfs of monthly data is slower than focusing on GSAT. Still much faster than an ESM, and once the recipes are set, very parallelizable. But for tractability, we are just going to stitch the gridded netcdfs for one generated ensemble member.

```{python}
#### python block ####
# Just do one realization
recipe = my_recipes[my_recipes['stitching_id'] == 'RCP  Standard-RCP-4.5~hectorUI1~1'].reset_index(drop=True).copy()

```


We must specify a directory for the netcdfs to be saved to. We will use simply do it directly here.
On my computer, constructing these 4 netcdfs took ~2-3 minutes.

```{python}
#### python block ####
stitches.gridded_stitching(out_dir='data', rp =recipe)
```


# plot gridded

- we will do maps of all 4 variables at a specified month and year
- we will do a time series of each variable in a specific grid cell

```{r, echo=FALSE}
#### R block ####
# pull all netcdfs:
files <- list.files('data', pattern  = '.nc', full.names = TRUE)

#subset to just the ESM of interest:
files <- files[grepl(py$esm, files)]

# subset to just the Hector scenario read in above:
files <- files[grepl(hector_exp, files)]

print(files)
```



# python maps and time series

It's a lot easier to use python to load in the netcdfs and select a grid cell or a time slice, save those off as simpler data frames than a full netcdf, and plot in R. The reshaping set of code that we often do in R to work with netcdf data is just so much faster if we've pre-sliced the data to a single grid cell or single time step in python.



## tas 
TODO just make it a function
```{python}
#### python block ####
py_files = pd.DataFrame(r.files, columns = ['file_name'])

# load the netcdf:
f = py_files.loc[py_files['file_name'].str.contains('tas')].values[0]
tas_nc = xr.open_dataset(f[0])

# time slice map:
tas_time_slice = tas_nc.sel(time = r.map_time).copy()


# time series for single grid cell:
# lon and lat values for a grid cell near College Park, MD, home of JGCRI:
cp_lat = 38.9897
cp_lon = 180 + 76.9378 # this is probably not actually right 

# lat and lon coordinates closest
abslat = np.abs(tas_nc.lat - cp_lat)
abslon = np.abs(tas_nc.lon-cp_lon)
c = np.maximum(abslon, abslat)
([lon_loc], [lat_loc]) = np.where(c == np.min(c))
lon_grid = tas_nc.lon[lon_loc]
lat_grid = tas_nc.lat[lat_loc]

tas_grid_slice = tas_nc.sel(lon = lon_grid, lat=lat_grid,
                         time = slice('1850-01-01', '2099-12-31')).copy()
del(tas_nc)
```

Take a look at this data in R:
```{r}
#### R block ####
# map data:
print(py$tas_time_slice$tas)
tas_val <- py$tas_time_slice$tas$values
tas_lon <- py$tas_time_slice$lon$values
tas_lat <- py$tas_time_slice$lat$values
grid <- expand.grid(list(lon=tas_lon,lat=tas_lat))
tas <- cbind(grid, 
             value=t(matrix(aperm(tas_val,c(2,1)), 1,length(tas_lat)*length(tas_lon))))

as.character(py$tas_time_slice$time) %>%
  substr(.,  37, 43) ->
  tas_time
  
ggplot(tas, aes(x = lon, y = lat, color=value, fill = value)) + geom_tile() +
  ggtitle(paste(tas_time, 'tas (K)'))


# grid time series data:
print(py$tas_grid_slice$tas)
cp_tas <- data.frame(value = py$tas_grid_slice$tas$values)

data.frame(time = as.character(py$tas_grid_slice$time$values)) %>%
  separate(time, into = c('year', 'month', 'day'), sep = '-') %>%
  select(-day) %>%
  mutate(year = as.integer(year),
         month = as.integer(month),
         time_index = as.integer(row.names(.))) ->
  cp_time

ggplot(cbind(cp_time, cp_tas), aes(x = time_index, y = value)) + geom_line() +
  ggtitle('College Park monthly tas, 1850-2100 (K)') 

ggplot(cbind(cp_time, cp_tas) %>% filter(year >= 2021), aes(x = time_index, y = value)) + geom_line() +
  ggtitle('College Park monthly tas, 2021-2100 (K)') 

```





## pr
```{python}
py_files = pd.DataFrame(r.files, columns = ['file_name'])

# load the netcdf:
f = py_files.loc[py_files['file_name'].str.contains('pr')].values[0]
pr_nc = xr.open_dataset(f[0])

# time slice map:
pr_time_slice = pr_nc.sel(time = r.map_time).copy()


# time series for single grid cell:
# lon and lat values for a grid cell near College Park, MD, home of JGCRI:
cp_lat = 38.9897
cp_lon = 180 + 76.9378 # this is probably not actually right 

# lat and lon coordinates closest
abslat = np.abs(pr_nc.lat - cp_lat)
abslon = np.abs(pr_nc.lon-cp_lon)
c = np.maximum(abslon, abslat)
([lon_loc], [lat_loc]) = np.where(c == np.min(c))
lon_grid = pr_nc.lon[lon_loc]
lat_grid = pr_nc.lat[lat_loc]

pr_grid_slice = pr_nc.sel(lon = lon_grid, lat=lat_grid,
                         time = slice('1850-01-01', '2099-12-31')).copy()
del(pr_nc)
```

Take a look at this data in R:
```{r}
# map data:
print(py$pr_time_slice$pr)
pr_val <- py$pr_time_slice$pr$values
pr_lon <- py$pr_time_slice$lon$values
pr_lat <- py$pr_time_slice$lat$values
grid <- expand.grid(list(lon=pr_lon,lat=pr_lat))
pr <- cbind(grid, 
             value=t(matrix(aperm(pr_val,c(2,1)), 1,length(pr_lat)*length(pr_lon))))

as.character(py$pr_time_slice$time) %>%
  substr(.,  37, 43) ->
  pr_time
  
ggplot(pr, aes(x = lon, y = lat, color=value, fill = value)) + geom_tile() +
  ggtitle(paste(pr_time, 'pr (kg m-2 s-1)'))


# grid time series data:
print(py$pr_grid_slice$pr)
cp_pr <- data.frame(value = py$pr_grid_slice$pr$values)

data.frame(time = as.character(py$pr_grid_slice$time$values)) %>%
  separate(time, into = c('year', 'month', 'day'), sep = '-') %>%
  select(-day) %>%
  mutate(year = as.integer(year),
         month = as.integer(month),
         time_index = as.integer(row.names(.))) ->
  cp_time

ggplot(cbind(cp_time, cp_pr), aes(x = time_index, y = value)) + geom_line() +
  ggtitle('College Park monthly pr (kg m-2 s-1), 1850-2100') 

ggplot(cbind(cp_time, cp_pr) %>% filter(year >= 2021), aes(x = time_index, y = value)) + geom_line() +
  ggtitle('College Park monthly pr (kg m-2 s-1), 2021-2100') 

```

## hurs
```{python}
py_files = pd.DataFrame(r.files, columns = ['file_name'])

# load the netcdf:
f = py_files.loc[py_files['file_name'].str.contains('hurs')].values[0]
hurs_nc = xr.open_dataset(f[0])

# time slice map:
hurs_time_slice = hurs_nc.sel(time = r.map_time).copy()


# time series for single grid cell:
# lon and lat values for a grid cell near College Park, MD, home of JGCRI:
cp_lat = 38.9897
cp_lon = 180 + 76.9378 # this is probably not actually right 

# lat and lon coordinates closest
abslat = np.abs(hurs_nc.lat - cp_lat)
abslon = np.abs(hurs_nc.lon-cp_lon)
c = np.maximum(abslon, abslat)
([lon_loc], [lat_loc]) = np.where(c == np.min(c))
lon_grid = hurs_nc.lon[lon_loc]
lat_grid = hurs_nc.lat[lat_loc]

hurs_grid_slice = hurs_nc.sel(lon = lon_grid, lat=lat_grid,
                         time = slice('1850-01-01', '2099-12-31')).copy()
del(hurs_nc)
```

Take a look at this data in R:
```{r}
# map data:
print(py$hurs_time_slice$hurs)
hurs_val <- py$hurs_time_slice$hurs$values
hurs_lon <- py$hurs_time_slice$lon$values
hurs_lat <- py$hurs_time_slice$lat$values
grid <- expand.grid(list(lon=pr_lon,lat=pr_lat))
hurs <- cbind(grid, 
             value=t(matrix(aperm(hurs_val,c(2,1)), 1,length(hurs_lat)*length(hurs_lon))))

as.character(py$hurs_time_slice$time) %>%
  substr(.,  37, 43) ->
  hurs_time
  
ggplot(hurs, aes(x = lon, y = lat, color=value, fill = value)) + geom_tile() +
  ggtitle(paste(hurs_time, 'hurs (%)'))


# grid time series data:
print(py$hurs_grid_slice$hurs)
cp_hurs <- data.frame(value = py$hurs_grid_slice$hurs$values)

data.frame(time = as.character(py$hurs_grid_slice$time$values)) %>%
  separate(time, into = c('year', 'month', 'day'), sep = '-') %>%
  select(-day) %>%
  mutate(year = as.integer(year),
         month = as.integer(month),
         time_index = as.integer(row.names(.))) ->
  cp_time

ggplot(cbind(cp_time, cp_hurs), aes(x = time_index, y = value)) + geom_line() +
  ggtitle('College Park monthly hurs (%), 1850-2100') 

ggplot(cbind(cp_time, cp_hurs) %>% filter(year >= 2021), aes(x = time_index, y = value)) + geom_line() +
  ggtitle('College Park monthly hurs (%), 2021-2100') 

```

## psl

sea level pressure can't have a CP time series
```{python}
py_files = pd.DataFrame(r.files, columns = ['file_name'])

# load the netcdf:
f = py_files.loc[py_files['file_name'].str.contains('psl')].values[0]
psl_nc = xr.open_dataset(f[0])

# time slice map:
psl_time_slice = psl_nc.sel(time = r.map_time).copy()
del(psl_nc)

```

Take a look at this data in R:
```{r}
# map data:
print(py$psl_time_slice$psl)
psl_val <- py$psl_time_slice$psl$values
psl_lon <- py$psl_time_slice$lon$values
psl_lat <- py$psl_time_slice$lat$values
grid <- expand.grid(list(lon=psl_lon,lat=psl_lat))
psl <- cbind(grid, 
             value=t(matrix(aperm(psl_val,c(2,1)), 1,length(psl_lat)*length(psl_lon))))

as.character(py$psl_time_slice$time) %>%
  substr(.,  37, 43) ->
  pr_time
  
ggplot(psl, aes(x = lon, y = lat, color=value, fill = value)) + geom_tile() +
  ggtitle(paste(pr_time, 'psl (Pa)'))

```
