---
title: "stitches in R - setting up environment"
author: "ACS"
date: '2022-09-22'
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# IMPORTANT NOTES

- Necessary to avoid R variable names like `variable.name` since `.` in python is somewhat analogous to `$` in R. 

- There are a bunch of steps you'll have to do before knitting this notebook.


- only tested on mac

- you might want a couple finder windows open to `usr/local/bin` and your analogous `/Users/snyd535/.virtualenvs/` so that you can check things are getting installed in the right places, and for the virtual environment, further navigate into `/Users/snyd535/.virtualenvs/r-reticulate/bin` and `/Users/snyd535/.virtualenvs/r-reticulate/lib/python3.9/site-packages` - the last one will let you see progress on dependency information because:

- once those steps are followed and this notebook is ready to be knit to install `stitches` from github, do be aware that it has a lot of dependencies that come along for the ride so it takes like 5-10 minutes. In Rstudio, it just kind of looks like it is doing nothing.


- I honestly I don't really understand virtual environments and especially `reticulate` seems sensitive, so if you want to add something to it later, honestly probably safer to delete the virtual environment and recreate from scratch here with the additions you want.

# BEFORE KNITTING

With what I figured out by trial and error in `figure-out-reticulate-setup.Rmd`, these are the steps need to set things up so that this markdown's python chunks are executed with the right version of python in the right virtual environment. 

## Step 1 - Python install
 This is outside of Rstudio
- stitches needs python 3.9 (most used to running with 3.9.7 I think)
- downloaded https://www.python.org/downloads/release/python-397/ to make sure it is in `usr/local/bin` on mac (`usr/local/bin/python3.9` exists)


## Step 1b
To avoid an issue when you get to step 6 and try to knit this notebook, go to 

`Applications` and select the dropdown next to `Python 3.9`. Click the `Install.Certificates.command` 

`stitches` needs to be able to access pangeo, which is online, so the whole thing fails without this.

More details including screenshots at the top answer: https://stackoverflow.com/questions/68275857/urllib-error-urlerror-urlopen-error-ssl-certificate-verify-failed-certifica

## Step 2 - `reticulate` install

need `reticulate` to be able to run both python and r code in the same file.

standard `install.packages("reticulate")` 

`Do you want to install from sources the package which needs compilation? (Yes/no/cancel)`

- I had issues with yes and couldn't track down the errors returned
- no worked fine, so went that direction

- because I didn't install the compiled version, I can't execute python blocks in Rstudio like I would an R block without knitting. They only execute when I knit. I can test python code in the Rstudio console once everything is set up with `reticulate::repl_python()`

## Step 3 - `r-reticulate` virtual environment

I could not figure out how to force reticulate to use the right version of python and have access to python packages without creating a virtual environment. Also, I _barely_ understand virtual environments, so I'm sure there are redundancies. But I'm superstitious so once i got something that worked from scratch, I stopped.  The below code block must be executed in Rstudio before knitting to set things up:

```{r, eval=FALSE}
###############################################################################
# Force reticulate to use python 3.9 so that python3.9 file gets included when
# we create the virtual environment

Sys.setenv("RETICULATE_PYTHON" = "/usr/local/bin/python3.9")


library(reticulate)

# create a new environment 
virtualenv_create("r-reticulate")

detach("package:reticulate", unload = TRUE)

###############################################################################
# reticulate is weird about being initialized to only a single python so
# that's why we do a lot of `detach` calls.

# Now Force reticulate to look for virtual env. Python 
Sys.setenv("RETICULATE_PYTHON" = "~/.virtualenvs/r-reticulate/bin/python3.9")

library(reticulate)

# # indicate that we want to use a specific virtualenv
# This should force the virtual environment to default to 3.9
use_virtualenv("r-reticulate", required =TRUE)

# check what's loaded now:
print('===========================================')
print('progress 1')
print(py_config())

###############################################################################
```

## Step 4 - close Rstudio and reopen project

This will happen a couple more times at the end of most subsequent steps

## Step 5 - install pkgs to virtual environment

These actually come along for the ride when you install stitches but I tested with more standard packages first to try to figure things out, so I left this in. Model for adding packages to virtual environment in future.

Now I think the virtual environment is setup to default to 3.9 when we load in `reticulate` and tell it to use the virtual environment fresh, without specifying it via `Sys.setenv`:
```{r, eval = FALSE}
library(reticulate)

# # indicate that we want to use a specific virtualenv
# This should force the virtual environment to default to 3.9
use_virtualenv("r-reticulate", required =TRUE)

# check what's loaded now:
print('===========================================')
print('progress 2')
print(py_config())

###############################################################################

# install and import python packages to virtual environment
# numpy
# install 
virtualenv_install("r-reticulate", "numpy")
# import 
numpy <- import("numpy")

# pandas
# install 
virtualenv_install("r-reticulate", "pandas")
# import 
pandas <- import("pandas")


```

You are now setup and can begin using `reticulate` in Rmarkdown and call the packages we just installed. Close Rstudio, reopen, and knit the notebook to execute all of the following:

# Step 6 - Reopen and knit markdown from this point



## installing stitches from github

Following: 
- https://stackoverflow.com/questions/42198066/how-to-install-a-local-package-using-pip-in-python-console 


`os` is a package that I think comes with default python. We can use it to pip install stitches from github and by doing it here, I think/hope it will get installed into the virtual environment so that we can use it in Rmarkdown notebooks.

 There's no way this is the best way to handle this but I've had so many issues with getting `reticulate` to use the  right things from the right places, I am just trying to be  consistent and careful.
 

```{r}
print('===========================================')
print('progress 3')
```
 
## setup `reticulate`

```{r}
library(reticulate)

# # indicate that we want to use a specific virtualenv
use_virtualenv("r-reticulate", required =TRUE)
```

## install `stitches` to correct virtual environment
```{python}
import os
os.system("pip3 install git+https://github.com/JGCRI/stitches@dev#egg=stitches")
```


Now you should be able to call `stitches` functions (and functions from its required packages) from a python block because it's all in the virtual environment. 

Once the steps in this notebook have been followed, R block code at the start of a notebook with the following:
```{r, eval=FALSE}
library(reticulate)

# # indicate that we want to use a specific virtualenv
use_virtualenv("r-reticulate", required =TRUE)
```

Will be all you need to use `stitches` and other python code. 

# Tests

## quick `reticulate` test

quick block of code that will only successfully knit if the setup was done correctly
```{python}
import numpy as np
print(np.max([2,3]))
```


# Load Hector data
pulled a quick rcp4.5 from HectorUI
```{r}
hector_tgav <- read.csv('data/Hector-data-2022-09-22.csv', skip=3, stringsAsFactors = FALSE)
head(hector_tgav)
```

# Make sure we can work with it from a python block
This is also a quick back and forth of moving data between R and python blocks following:

- https://rstudio.github.io/reticulate/articles/r_markdown.html#calling-python-from-r
- https://rstudio.github.io/reticulate/articles/r_markdown.html#calling-r-from-python



```{python}
import numpy as np
import pandas as pd

tgav_df = pd.DataFrame(r.hector_tgav)
print(np.max(tgav_df[['value']]))

```

and we can compare to the R evaluation
```{r}
# The hector data we read straight in to R
print(max(hector_tgav$value))

# R operations on the python data
print(max(py$tgav_df$value))
```

# make sure can import stitches and call a function from it
 and have it still work and be nice
This is a good function to test a `stitches` installation with - no arguments needed, just a list of data on pangeo, implicitly tests that can connect to pangeo. 

```{python}
import stitches
pangeo_table = stitches.fx_pangeo.fetch_pangeo_table()
print(pangeo_table.head())
```

# Then probably close Rstudio and reopen before running other notebooks
